---
title: "An Analysis of Obesity Prevelance in Scotland (2013-2016)"
author: "Group 1 "
execute: 
  echo: false
  eval: true
number-sections: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(AICcmodavg)
library(readr)
library(ggplot2)
library(ggfortify)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
library(stringr)
library(sjPlot)
library(knitr)
```

```{r}
#| echo: false
#| warning: false
#| message: false
obesity <- read.csv("DAProject3.csv")
```

# Introduction {#sec-intro}

Obesity occurs when an individual's energy intake from food and drink is higher than the amount of energy needed over a period of time, and is considered a serious public health concern. Data for this report has been gathered from the Scottish Health Surveys from 2013-2016. The data includes the binary variable, `Obese`, which indicates whether an individual classes themselves as obese (Yes), or not obese (No), and the following other lifestyle and socio-economic factors:

-   `AgeGroup` - The age range of the individual (a categorical variable with seven different age categories).

-   `Sex` - The sex of an individual (a binary variable with categories Male or Female).

-   `Employment` - The individual's employment status (a categorical variable with seven different employment categories).

-   `Veg` - Indicates whether the individual consumes the recommended daily intake of vegetables (a binary variable with categories Yes or No).

-   `Fruit` - Indicates whether the individual consumes the recommended daily intake of fruit (a binary variable with categories Yes or No).

-   `Year` - The year in which the Scottish Health Survey was conducted (a categorical variable with categories corresponding to years 2013, 2014, 2015 and 2016).

The main focus of this report will be to examine the prevalence of obesity in Scotland over the given years and to examine differences in obesity by age, sex and the other lifestyle and socio-economic factors.

@sec-exploranalysis consists of an exploratory analysis of the obesity data and explores the potential relationship between factors. @sec-formalanalysis contains the results from fitting a logistic regression model to the data, as well as the assessment of the model assumptions. Concluding remarks are given in @sec-conclusion.

# Exploratory Analysis {#sec-exploranalysis}

@fig-bar1 displays the relationship between variables `Obesity` and `Year` revealing the prevalence of obesity in Scotland over the given years of the Scottish Health Survey. From @fig-bar1, we can see that, on average, there was not much change in the recorded number of obese individuals in Scotland by year from 2013 to 2016. We observe a slightly smaller count in 2016 compared to the previous three years, but it appears that `Year` is not a significant explanatory variable.

The proportion of no obesity has remained way above the proportion for obesity over the given years.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-bar1
#| fig-cap: Relationship between Obesity and Year.
#| fig-align: center
#| fig-height: 3.5
ggplot(obesity, aes(x = Year, group = Obese)) + 
  geom_bar(aes(y = after_stat(prop), fill = Obese), stat = "count", position = "dodge") +
  labs(y = "Proportion", fill = "Obese")
```

In @fig-bar2 we observe the relationships between `Obesity` and `AgeGroup`, and `Obesity` and `Sex`. It is revealed that the obesity proportion is highest in the 55-64 age group and decreases as age decreases, as well as for an increase in age. The lowest obesity proportion is counted in the 16-24 age group.

We can also see in @fig-bar2 that obesity appears slightly higher in females than in males.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-bar2
#| fig-cap: Obesity by Age Group and Sex.
#| fig-align: center
#| fig-height: 3.5
obesity_sub <- obesity |> dplyr::filter(Obese=="Yes")

p1 <- ggplot(obesity_sub, aes(x = AgeGroup, group = Obese)) + 
      geom_bar(aes(y = after_stat(prop)), fill = "turquoise3", stat = "count", position = "dodge") +
      labs(y = "Proportion", fill = "Obese")

p2 <- ggplot(obesity, aes(x = Sex, group = Obese)) + 
  geom_bar(aes(y = after_stat(prop), fill = Obese), stat = "count", position = "dodge") +
  labs(y = "Proportion", fill = "Obese")


p1+p2+plot_layout(ncol = 2)
```

We also observe the relationship between `Obesity` and different lifestyle factors (whether or not the individual consumes the recommended daily intake of fruit or vegetables) in @fig-bar3. Obesity seems to be more prevalent in those who did consume the recommended daily intake of fruit or veg compared to those who didn't. However, there is little to no difference between the proportion of obese people and not obese people, so these variables (`Fruit` in particular) seem insignificant at this stage.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-bar3
#| fig-cap: Relationship between Obesity and Lifestyle Factors.
#| fig-align: center
#| fig-height: 3.5

# fruit
f <- ggplot(obesity, aes(x = Fruit, group = Obese)) + 
  geom_bar(aes(y = after_stat(prop), fill = Obese), stat = "count", position = "dodge") +
  labs(y = "Proportion", fill = "Obese")

# veg
v <- ggplot(obesity, aes(x = Veg, group = Obese)) + 
  geom_bar(aes(y = after_stat(prop), fill = Obese), stat = "count", position = "dodge") +
  labs(y = "Proportion", fill = "Obese")

f+v+plot_layout(ncol = 2)
```

The relationship between `Obesity` and `Employment` is revealed in @fig-bar4. The obesity proportion is clearly highest in employed individuals, followed by retirees, while the other factors appear to have very low proportions of obesity.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-bar4
#| fig-cap: Relationship between Obesity and Socio-economic Status. 
#| fig-align: center
#| fig-height: 3.5

ggplot(obesity_sub, aes(x = Employment, group = Obese)) + 
  geom_bar(aes(y = after_stat(prop), fill = Obese), fill = "turquoise3", stat = "count", position = "dodge") +
  labs(y = "Proportion") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))

```

Finally @fig-bar5 is plotted to see if there is any relationship between `AgeGroup` and `Sex`. Initial inspection of @fig-bar5 highlights that there does seem to be some difference between proportions of `Sex` by `AgeGroup` between some of the age groups.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-bar5
#| fig-cap: Relationship between Age and Sex. 
#| fig-align: center
#| fig-height: 3.5
ggplot(obesity, aes(x = AgeGroup, y = after_stat(prop), fill = Sex, group = Sex)) + 
  geom_bar(position = "dodge", stat = "count") +
  labs(y = "Proportion") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))

```

# Formal Analysis {#sec-formalanalysis}

```{r}
#| echo: false
#| warning: false
#| message: false
# change to factors
obesity$AgeGroup <- as.factor(obesity$AgeGroup)
obesity$Year <- as.factor(obesity$Year)
obesity$Sex <- as.factor(obesity$Sex)
obesity$Employment <- as.factor(obesity$Employment)
obesity$Veg <- as.factor(obesity$Veg)
obesity$Fruit <- as.factor(obesity$Fruit)
obesity$Obese <- as.factor(obesity$Obese)
```

To analyse the prevalence of obesity over the given years, we produce a logistic regression model for `Obesity` by `Year`. The large p-values shown in @fig-GLMyear match our assumptions from @fig-bar1: we can conclude that year has no significant effect on obesity in Scotland.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-GLMyear
#| fig-cap: GLM with Year as explanatory variable.
year.model <- glm(Obese ~ Year, data=obesity, family = "binomial")
coefficents.year <- year.model |> 
  summary() |> 
  coef()
knitr::kable(coefficents.year)
```

Now we go ahead and fit a model in order to analyse the relationship between obesity and our other variables. The `Year` and `Fruit` variables are not included in our initial fit of the model: our exploratory analysis concluded that these variables do not appear to impact obesity. Also, there does not appear to be any interaction between sex and employment so we also remove this interaction from the final model. The `StepAIC` function from the `MASS` package is then used to fit the model with the lowest AIC, the output of which is displayed in @fig-GLMfull.

Note that the baseline category for our response of obese is "No". This means that the estimates for our fitted logistic regression model are for changes in log-odds for people that answered "Yes" to a self evaluation of being obese against the baseline of a self-evaluation of "No".

```{r}
#| echo: false
#| warning: false
#| message: false
obesity.sub2 <- obesity |> dplyr::select(Obese,Veg,Sex,Employment,AgeGroup)
init.model <- glm(Obese~.^2-Sex:Employment , data = obesity.sub2, family = "binomial")
final.model <- stepAIC(init.model,direction = "both", trace = FALSE)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-GLMfull
#| fig-cap: Summary of regression coefficients for the final model
reg.table <- final.model |> summary() |> coef()
reg.table <- as.data.frame(reg.table)
mod.final.lower <- reg.table$Estimate - (1.96 * reg.table$`Std. Error`)
mod.final.upper <- reg.table$Estimate + (1.96 * reg.table$`Std. Error`)
reg.table <- reg.table |> mutate(Lower.Bound = mod.final.lower, Upper.Bound = mod.final.upper)
knitr::kable(reg.table)
```

@fig-GLMfull shows each of the parameters that are included in the final model. The model selected by the `StepAIC` function includes `Veg` , `Sex` , `Employment` , `AgeGroup` and an interaction between `Sex` and `Agegroup` . The full fitted model is formally written below:

```{=tex}
\begin{equation*}
  \begin{aligned}
    \ln\left(\frac{p}{1-p}\right) &= -0.165\cdot\mathbb{I}_\text{Veg} -0.569\cdot\mathbb{I}_\text{Male}  -0.291\cdot\mathbb{I}_\text{Full-time Employment} \\
    &-0.032\cdot\mathbb{I}_\text{In Paid Employment/Self Employed/Govt Training} 
+0.375\cdot\mathbb{I}_\text{Looking after home/family} \\

&+ 0.248\cdot\mathbb{I}_\text{Looking for/Intending to look for work} +0.443\cdot\mathbb{I}_\text{Perm unable to work} + 0.049\cdot\mathbb{I}_\text{Retired} \\ 

&+0.308\cdot\mathbb{I}_\text{25-34} +0.511\cdot\mathbb{I}_\text{35-44} +0.632\cdot\mathbb{I}_\text{45-54} + 0.618\cdot\mathbb{I}_\text{55-64} + 0.775\cdot\mathbb{I}_\text{65-74} \\ 

&+0.562\cdot\mathbb{I}_\text{75+}+ 0.312\cdot\mathbb{I}_\text{Male and 25-34} + 0.348\cdot\mathbb{I}_\text{Male and 35-44} + 0.500\cdot\mathbb{I}_\text{Male and 45-54} \\


&+ 0.723\cdot\mathbb{I}_\text{Male and 55-64} + 0.571\cdot\mathbb{I}_\text{Male and 65-74} + 0.237\cdot\mathbb{I}_\text{Male and 75+} \\

  \end{aligned}
\end{equation*}
```
Where, $\mathbb{I}$ is an indicator function and $p = \text{Prob}(\text{Obese})$. From @eq-fittedmodel it highlights how the log-odds of an individual self-evaluating as obese changes by the various lifestyle and socio-economic factors. This will then give us estimates for the log-odds, and therefore odds for each individual in the survey.

We now wish to examine ranges of plausible log-odds of obesity for each of the predictors in @eq-fittedmodel computing 95% confidence intervals. Note that these confidence intervals will also highlight significance of the parameters, this will give the same result as the conclusion reached in @fig-GLMfull.

@fig-CI shows the 95% confidence intervals for each regression parameter included in the fitted model. For example, we are 95% confident that the log-odds of an individual being obese is 0.078 to 0.252 lower if the individual eats vegetables.

Finally, to see if our model predicts as we would expect it to we produce plots of the probability of an individual self-evaluation as obese by the various lifestyle and socio-conomic factors. @fig-probs1 shows these produced plots. From the plots it is clear that the probabilities of an individual being obese clearly vary by these factors and in fact, these probabilities matched what we saw from our plots in @sec-exploranalysis.

##  Assessing model assumptions

# Conclusion {#sec-conclusion}
