---
title: "Analysis of Obesiety Prevelance in Scotland (2013-2016)"
author: "Group 1 "
execute: 
  echo: false
  eval: true
number-sections: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(AICcmodavg)
library(readr)
library(ggplot2)
library(ggfortify)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
library(stringr)
```

```{r}
#| echo: false
#| warning: false
#| message: false
obesity <- read.csv("DAProject3.csv")
```

# Introduction {#intro}

# Exploratory Analysis {#exploranalysis}

@fig-bar1 displays the relationship between variables `Obesity` and `Year` revealing the prevalence of obesity in Scotland over the given years of the Scottish Health Survey. From @fig-bar1, we can see that, on average, there was not much change in the recorded number of obese individuals in Scotland by year from 2013 to 2016. We observe a slightly smaller count in 2016 compared to the previous three years, but not much difference in the data from 2013 to 2015.

The count for no obesity has remained way above the count for obesity over the given years.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-bar1
#| fig-cap: Relationship between Obesity and Year.
#| fig-align: center
#| fig-height: 3.5
ggplot(obesity, aes(x = Year, fill=Obese)) + geom_bar(position = 'dodge') +
  labs(y = "Count")
```

In @fig-bar2 we observe the relationships between `Obesity` and `AgeGroup`, and `Obesity` and `Sex`. It is revealed that the obesity count is highest in the 55-64 age group and decreases as age decreases, as well as for an increase in age. The lowest obesity number is counted in the 16-24 age group.

We can also see in @fig-bar2 that the obesity count appears slightly higher in females than in males.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-bar2
#| fig-cap: Obesity by Age Group and Sex.
#| fig-align: center
#| fig-height: 3.5
obesity_sub <- obesity |> dplyr::filter(Obese=="Yes")
p1 <- ggplot(obesity_sub, aes(x = AgeGroup)) + geom_bar(fill = "turquoise3") +
      labs(y = "Count")

p2 <- ggplot(obesity, aes(x = Sex, fill = Obese)) + geom_bar(position = 'dodge') + labs(y = "Count")

p1+p2+plot_layout(ncol = 2)
```

We also observe the relationship between `Obesity` and different lifestyle factors (whether or not the individual consumes the recommended daily intake of fruit or vegetables) in @fig-bar3. Obesity seems to be more prevalent in those who did consume the recommended daily intake of fruit or veg compared to those who didn't.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-bar3
#| fig-cap: Relationship between Obesity and Lifestyle Factors.
#| fig-align: center
#| fig-height: 3.5
p1 <- ggplot(obesity, aes(x = Fruit, fill=Obese)) + geom_bar(position = 'dodge') +
  labs(y = "Count")

p2 <- ggplot(obesity, aes(x = Veg, fill=Obese)) + geom_bar(position = 'dodge') +
  labs(y = "Count")

p1+p2+plot_layout(ncol = 2)
```

Finally, the relationship between `Obesity` and `Employment` is revealed in @fig-bar4. The obesity count is clearly highest in employed individuals, followed by retirees, while the other factors appear to have very low obesity counts.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-bar4
#| fig-cap: Relationship between Obesity and Socio-economic Status. 
#| fig-align: center
#| fig-height: 3.5
ggplot(obesity_sub, aes(x = Employment)) + 
         geom_bar(fill = "turquoise3") +
  labs(y = "Count") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))
```

# Formal Analysis {#formalanalysis}

```{r}
#| echo: false
#| warning: false
#| message: false

# change to factors
obesity$AgeGroup <- as.factor(obesity$AgeGroup)
obesity$Year <- as.factor(obesity$Year)
obesity$Sex <- as.factor(obesity$Sex)
obesity$Employment <- as.factor(obesity$Employment)
obesity$Veg <- as.factor(obesity$Veg)
obesity$Fruit <- as.factor(obesity$Fruit)
obesity$Obese <- as.factor(obesity$Obese)
```

To answer our first question of interest we produce a logistic regression model for `Obesity` by `Year`. The large p-values shown in the following output match our assumptions from @fig-bar1: we can conclude that year has no significant effect on obesity in Scotland.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-output

summary(glm(Obese ~ Year, data=obesity, family = "binomial"))
```

Full model:

```{r}

final.model <- glm(Obese ~ Veg + Sex * AgeGroup + Sex * Employment, family = 'binomial', data = obesity)

summary(final.model)
```

# Conclusion {#conclusion}
